---
title: "MATH 216 Homework 4"
author: "Andrew Holtz"
output: html_document
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# For data manipulation and visualization
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(knitr))
# For US county and state maps
suppressPackageStartupMessages(library(maps))
# For loading in shapefiles
suppressPackageStartupMessages(library(rgdal))
suppressPackageStartupMessages(library(maptools))
# For interactive maps
suppressPackageStartupMessages(library(leaflet))
suppressPackageStartupMessages(library(stringr))
suppressPackageStartupMessages(library(plotly))
suppressPackageStartupMessages(library(pander))
```

## Admistrative:

Please indicate

* Who you collaborated with: Mohamed Hussein
* Roughly how much time you spent on this HW: 8 hours
* What gave you the most trouble: Figuring out what data set to download and 
then some how merge this with the VT data- Mohamed helped me with this. 
* Any comments you have: Maps are cool



## Question 1:

```{r, cache=TRUE, echo=FALSE}
# Load state and county map of US in 2010 from the maps package and convert them
# to data frames so that we can ggplot them.
US_state <- map_data("state") %>% 
  tbl_df()
US_county <- map_data("county") %>% 
  tbl_df()

COUNTY <-read.csv("Data/county.csv")

```

###Choropleth Map of US Voter Behavior in 2000

Download the results of the 2000 election from the [School of Public Affairs at 
American University in DC](http://www.american.edu/spa/ccps/Data-Sets.cfm) and 
create a map involving only the lower 48 states that show voter behavior at a 
**county** level. To keep things simple lets only consider George W. Bush, Al
Gore, and Ralph Nader. Your write-up must include:

1. A [choropleth map](https://en.wikipedia.org/wiki/Choropleth_map) where
    * Counties are filled with red when they tend to favor Bush
    * Counties are filled with white when they tend to be split
    * Counties are filled with blue when they tend to favor Gore and Nader
2. An answer to the following question: which states exhibit the greatest
**within state heterogeneity** in voting? Come up with a mathematical
justification.


### Hints

* Consider `scale_fill_gradient2(name="", low="blue", high="red", mid="white")` 
for the appropriate "mid" point.  See the ggplot2 webpage for [this
command](http://docs.ggplot2.org/0.9.3.1/scale_gradient2.html) for inspiration.
* I believe the county map is from 2010, whereas the election data is from 2000,
as such certain counties will not match up. The number is not that large.
* [Counties in
Virginia](http://en.wikipedia.org/wiki/List_of_counties_in_Virginia)


### Code Hints

This function eliminates all non-alphanumeric characters and spaces and converts
all text to lower case:

```{r, echo = FALSE, warning = FALSE}
clean_text <- function(text){
  text <- gsub("[^[:alnum:]]", "", text)
  text <- gsub(" ", "", text)
  text <- tolower(text)
  return(text)
}
```

For the following `ggplot`, see how I didn't define `data` nor any `aes`thetics
in the base `ggplot`, but only within the respective `geom`'s. This is a nice
trick that allows you to have different `data` and different `aes` apply to
different `geom`'s; in this case state-level vs county-level data.

```{r, echo = FALSE, warning = FALSE}

pgore <- as.numeric(as.character(COUNTY$PGORE))
pnader <- as.numeric(as.character(COUNTY$PNADER))
COUNTY$PBUSH <- as.numeric(as.character(COUNTY$PBUSH))
COUNTY$PGORE <- as.numeric(as.character(COUNTY$PGORE))
COUNTY$PNADER <- as.numeric(as.character(COUNTY$PNADER))
COUNTY$POTHER <- as.numeric(as.character(COUNTY$POTHER))
COUNTY <- mutate(COUNTY, state = clean_text(STATE))

st_co <- COUNTY %>% 
  mutate(PGO_NA = pgore + pnader) %>% 
  mutate(state = clean_text(STATE)) %>% 
  mutate(county = clean_text(COUNTY))

st_co <- st_co %>% 
  mutate(state = str_replace_all(state, fixed(" "),"")) %>% 
  mutate(county = str_replace_all(county, fixed(" "),"")) %>% 
  select(state, county, PBUSH, PGO_NA)

US_county <- US_county %>% 
  mutate(region = str_replace_all(region, fixed(" "),"")) %>% 
  mutate(subregion = str_replace_all(subregion, fixed(" "),""))
  

stco_coor <- right_join(US_county, st_co, 
                        by = c("region" = "state", "subregion" = "county"))


ggplot(data=NULL) +
  geom_path(data=US_county, aes(x=long, y=lat, group=group), col="black", size=0.05) +
  geom_path(data=US_state, aes(x=long, y=lat, group=group), col="black", size=0.1) +
  geom_polygon(data = stco_coor, aes(x=long, y=lat, group=group, fill = PBUSH)) +
  coord_map() +
  scale_fill_gradientn(name="Election Result", 
                       colours = c("#261758", "white", "#801515"),
                       values = (c(0,0.499, 0.5, 0.5001, 1.0)))

#Finding within state heterogeneity

hetero <- COUNTY %>% 
  filter(!is.na(PBUSH)) %>% 
  select(state, PBUSH, PGORE, PNADER, POTHER) %>% 
  group_by(state) %>% 
  summarise(b_avg = mean(PBUSH), g_avg = mean(PGORE), n_avg = mean(PNADER), o_avg = 
              mean(POTHER)) %>% 
  arrange(b_avg, o_avg)

hetero <- hetero %>% 
  mutate(sum_bg = b_avg + g_avg) %>% 
  mutate(sum_bgn = sum_bg + n_avg) %>% 
  arrange(sum_bgn) %>% 
  top_n(5,-sum_bgn)


panderOptions("table.style", "rmarkdown")
pander(hetero, split.table = Inf)

```


### Write-Up
As you can see from this map, the coasts prodominately voted democrat. The
counties that voted for Bush are seen to be in the middle of the country. This
also follows city lines. The counties with large cities (also the smaller of 
counties) voted democratic, while the more rural, larger counties, voted for
Bush. 

You can see from the hetero pander table that north dakota and alaska are the 
most heterogeneous states in the country. You can see this based off the sum_bg
which sums the average proprotion of Bush voters with the average proportion of
Gore voters. Here, Alaska takes first place, meaning that voters in Alaska were 
more likely to vote for candidates besides the democratic and republican nominee
compared to voters in any other state. It is interesting, however, that when you
factor in Nader (protrayed by sum_bgn), you see that North Dakota is actually 
the most heterogeneous with Alaska as a close second. This means that North 
Dakota voters were more likely compared to voters in any other state to vote for candidates besides Bush, Gore, and Nader. This is also shown true by the 
proportion "other" column (o_avg). 

```{r, fig.width=12, fig.height=6}

```

## Question 2:

In this question, you must make an interactive "Single File" Shiny app that uses
Leaflet. For all 184 census tracts in VT in the 2010 census, present information
on the proportion of the population that is either

* White
* African-American
* Hispanic
* Asian and Pacific Island
* Other, including those of two or more races

Use [Social Explorer](http://www.socialexplorer.com/) to get census data. I did
a demo of this in class. If you don't remember how or are stuck, please speak to
me or get help from your peers. Do not submit copies of the same file.

There should be some mechanism in your Shiny app that allows one the user to 
toggle between the different ethnic groups.


### Loading Shapefile Data

Here is some starter code:

```{r, eval = TRUE, echo = TRUE}

shapefile_name <- paste(getwd(), "/VT_census_tracts/tl_2015_50_tract.shp", sep="")

VT <- readOGR(shapefile_name, layer = "tl_2015_50_tract", verbose = FALSE)

census <- read.csv("Data/R11168779_SL140.csv")
hispanic <- read.csv("Data/hispanic.csv")


census <- census %>% select(Geo_NAME, Geo_TRACT, Geo_FIPS, SE_T054_001,	SE_T054_002,
                            SE_T054_003,	SE_T054_004,	SE_T054_005,	SE_T054_006,
                            SE_T054_007,	SE_T054_008)
  


census <- mutate(census, hisp = hispanic$SE_T055_010)

# Rename columns
names(census)[4:12] <- c("tot_pop","white_pop","black_pop","nativ","asian","hawaii",
                        "other", "mixed", "hisp")

# proportions of each race in VT by total pop
census <- census %>% 
  mutate(
  prop_white = white_pop/tot_pop,
  prop_black = black_pop/tot_pop,
  prop_nativ = nativ/tot_pop, 
  prop_asian = asian/tot_pop, 
  prop_hawaii = hawaii/tot_pop, 
  prop_other = other/tot_pop,
  prop_mixed = mixed/tot_pop,
  prop_hisp = hisp/tot_pop)
  
#Prop_white
leaflet(VT) %>%
  addPolygons(
    stroke = FALSE, fillOpacity = 0.8, smoothFactor = 0.5,
    color = ~colorQuantile("Blues", census$prop_white)(census$prop_white)) %>% 
  addTiles() %>% 
  addPolylines(color="black", weight=1) %>% 
  addLegend("bottomright", values = ~census$prop_white,
    title = "Proportion White of Total Pop",
    labFormat = labelFormat(prefix = ""),
    opacity = 1, pal = colorNumeric("Blues", census$prop_white))

#Mo showed me how to do this

#Prop_black
leaflet(VT) %>%
  addPolygons(
    stroke = FALSE, fillOpacity = 0.8, smoothFactor = 0.5,
    color = ~colorQuantile("BuGn", census$prop_black)(census$prop_black)) %>% 
  addTiles() %>% 
  addPolylines(color="black", weight=1) %>% 
  addLegend("bottomright", values = ~census$prop_black,
    title = "Proportion Black of Total Pop",
    labFormat = labelFormat(prefix = ""),
    opacity = 1, pal = colorNumeric("BuGn", census$prop_black))

#Prop_asian
leaflet(VT) %>%
  addPolygons(
    stroke = FALSE, fillOpacity = 0.8, smoothFactor = 0.5,
    color = ~colorQuantile("RdPu", census$prop_asian)(census$prop_asian)) %>% 
  addTiles() %>% 
  addPolylines(color="black", weight=1) %>% 
  addLegend("bottomright", values = ~census$prop_asian,
    title = "Proportion Asian of Total Pop",
    labFormat = labelFormat(prefix = ""),
    opacity = 1, pal = colorNumeric("RdPu", census$prop_asian))

#Prop_hispanic
leaflet(VT) %>%
  addPolygons(
    stroke = FALSE, fillOpacity = 0.8, smoothFactor = 0.5,
    color = ~colorQuantile("Oranges", census$prop_hisp)(census$prop_hisp)) %>% 
  addTiles() %>% 
  addMarkers(lng=-73.2540, lat=44.1673, popup="Vergennes") %>% 
  addPolylines(color="black", weight=1) %>% 
  addLegend("bottomright", values = ~census$prop_hisp,
    title = "Proportion Hispanic of Total Pop",
    labFormat = labelFormat(prefix = ""),
    opacity = 1, pal = colorNumeric("Oranges", census$prop_hisp))

#two or more races
leaflet(VT) %>%
  addPolygons(
    stroke = FALSE, fillOpacity = 0.8, smoothFactor = 0.5,
    color = ~colorQuantile("Reds", census$prop_other)(census$prop_other)) %>% 
  addTiles() %>% 
  addPolylines(color="black", weight=1) %>% 
  addLegend("bottomright", values = ~census$prop_other,
    title = "Proportion of >2 races of Total Pop",
    labFormat = labelFormat(prefix = ""),
    opacity = 1, pal = colorNumeric("Reds", census$prop_other))

```


### Write-Up

Overall, one can see that Vermont is predominately white. That is not to say,
however, that there aren't some pockets of diversity. The Northwest corner of 
Vermont seems to have a lot of diversity. You can see high proportions of black
people, asian, and hispanic in this region of Vermont. If you look at this 
region on the White proportion map, you can see that it has low proportions of 
white compared to the rest of Vermont. Between black, asian, hispanics, and 
other, you can see that blacks and asians are very similarly represented as
minority ethnic groups in Vermont. The highest for both ethnicities / race 
is ~8-9%. Hispanics are represented much less with the census tract 
with largest hispanic population at ~4%. There are very few people living in 
Vermont who identify as having two or more races. The census tract with the 
largest population of people with two or more is about 1.3%. Interestingly,
I know that there is a large proportion of hispanics that live in Vergennes 
because I have friends who work at an Open Door Clinic there. I know that they
often interact with patients who only speak Spanish. The map of proportion of 
Hispanic people matches with this observation. If one were to zoom in at
Vergennes, one can see that it has a census tract with one of the highest
proportions of hispanics. 


